---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 1"
author: "Mark Fairchild"
date: "8 Oct 2021"
output: html_notebook
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

type in View(dennys) into the lower shell area to get data frame
```{r load-data, message = FALSE}
states <- read_csv("U:/data Science 241/states.csv")
```

```{r load_laquinta}
dsbox::laquinta -> laquinta
laquinta=laquinta
```

### Exercise 1

What are the dimensions of the Denny’s dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) What does each row in the dataset represent? What are the variables?
```{r dimensions_of_dennys}
nrow(dennys)
ncol(dennys)
```
The dimentions are 6 columns with 1643 rows each
each row represents a dennys store location
each column represents an aspect of that location

### Exercise 2

What are the dimensions of the La Quinta’s dataset? What does each row in the dataset represent? What are the variables?
```{r dimensions_of_laquinta}
nrow(laquinta)
ncol(laquinta)
```
The dimentions are 6 columns with 909 rows each
each row represents a laquinta store location
each column represents an aspect of that location

### Exercise 3
Take a look at the websites that the data come from (linked above). Are there any La Quinta’s locations outside of the US? If so, which countries? What about Denny’s?
...
La Quinta has locations outside of the US. The countries include Canada, Mexico, New Zealand, Honduras, Turkey, the UAE, Chile, and Colombia.

Dennys appears to have no locations outside of the US.
### Exercise 4
Now take a look at the data. What would be some ways of determining whether or not either establishment has any locations outside the US using just the data (and not the websites). Don’t worry about whether you know how to implement this, just brainstorm some ideas. Write down at least one as your answer, but you’re welcomed to write down a few options too.
...
One way would be to look at the lattitude and longitute. You could filter the rows so that only observations that were not (!) within the lattitude and longitute of the US were shown.
### Exercise 5
Find the Denny’s locations that are outside the US, if any. To do so, filter the Denny’s locations for observations where state is not in states$abbreviation. The code for this is given below. Note that the %in% operator matches the states listed in the state variable to those listed in states$abbreviation. The ! operator means not. Are there any Denny’s locations outside the US?
...
```{r dennys_locations_outside_us}
dennys %>%
  filter(!(state %in% states$abbreviation))
```
There are no Denny's locations outside the US. The code gave us zero results!

### Exercise 6
Add a country variable to the Denny’s dataset and set all observations equal to "United States". Remember, you can use the mutate function for adding a variable. Make sure to save the result of this as dennys again so that the stored data frame contains the new variable going forward.

```{r add_column}
dennys2 <- dennys %>%
  mutate(country = "United States")
```

...

### Exercise 7
Find the La Quinta locations that are outside the US, and figure out which country they are in. This might require some googling. Take notes, you will need to use this information in the next exercise.
...
```{r find_state_abbreviations_for_locations_in_Mexico}
# non_us_locations_of_laquinta <- laquinta %>%
#   filter(!(state %in% states$abbreviation)) %>%
#    filter(state != "ON", 
#           state != "BC", 
#           state != "ANT", 
#           state != "FM")
```

```{r}
 laquinta %>%
   filter(!(state %in% states$abbreviation))
```

In order of appearance, the 14 non US locations are in the following countries
1.Mexico 2.Mexico 3.Mexico 4.Mexico 5.Columbia 6.Mexico 7.Mexico 8.Canada 9.Mexico 10.Mexico 11.Mexico 12.Mexico 13.Honduras 14.Canada
### Exercise 8
Add a country variable to the La Quinta dataset. Use the case_when function to populate this variable. You’ll need to refer to your notes from Exercise 7 about which country the non-US locations are in. Here is some starter code to get you going:
...
```{r}
laquinta <- laquinta %>%
  mutate(country = case_when(
    state %in% state.abb     ~ "United States",
    state %in% c("ON", "BC") ~ "Canada",
    state == "ANT"           ~ "Colombia",
    state == "FM"            ~ "Honduras",
    state %in% c("AG", "QR", "CH", "NL", "VE", "PU", "SL") ~ "Mexico"

  ))
```

```{r}
laquinta <- laquinta %>%
  filter(country == "United States")
```

### Exercise 9
Which states have the most and fewest Denny’s locations? What about La Quinta? Is this surprising? Why or why not?
Next, let’s calculate which states have the most Denny’s locations per thousand square miles. This requires joining information from the frequency tables you created in Exercise 8 with information from the states data frame.

First, we count how many observations are in each state, which will give us a data frame with two variables: state and n. Then, we join this data frame with the states data frame. However note that the variables in the states data frame that has the two-letter abbreviations is called abbreviation. So when we’re joining the two data frames we specify that the state variable from the Denny’s data should be matched by the abbreviation variable from the states data:
...
```{r number_of_dennys_locations_by_state}
dennys3 <- dennys2 %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))
```
Delaware has the least dennys locations. This is not suprising since it is a small state. Texas has the most locations. Again this is not suprising since it is the second biggest mainland state.

...

```{r}
laquinta2 <- laquinta %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))
```
Maine has the least laquinta locations. I find this a little suprising since there is definitely some tourist attractions in Maine. Once again Texas has the most locations. This is still unsuprising for the same reasons as in the case of Dennys.

### Exercise 10
Which states have the most Denny’s locations per thousand square miles? What about La Quinta?

```{r dennys_per_thousand_miles}
dennys_per_sqm <- dennys3 %>%
mutate(location_per_sqm = n/(area/1000))
```


```{r laquinta_per_thousand_miles}
laquintas_per_sqm <- laquinta2 %>%
mutate(location_per_sqm = n/(area/1000))
```
Both La Quinta and Denny's have the most locations per square mile in the state of Rhode Ilsand. 

...
Next, we put the two datasets together into a single data frame. However before we do so, we need to add an identifier variable. We’ll call this establishment and set the value to "Denny's" and "La Quinta" for the dennys and laquinta data frames, respectively.

```{r add_establishment_variable_to_dennys}
dennys2 <- dennys2 %>%
  mutate(establishment = "Denny's")
```

```{r add_establishment_variable_to_laquinta}
laquinta <- laquinta %>%
  mutate(establishment = "La Quinta")
```
  
  Since the two data frames have the same columns, we can easily bind them with the bind_rows function
  
```{r bind_dataframes}
dn_lq <- bind_rows(dennys, laquinta)
```

We can plot the locations of the two establishments using a scatter plot, and color the points by the establishment type. Note that the latitude is plotted on the x-axis and the longitude on the y-axis

```{r plot_locations}
ggplot(dn_lq, mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point()
```

...

### Exercise 11
Filter the data for observations in North Carolina only, and recreate the plot. You should also adjust the transparency of the points, by setting the alpha level, so that it’s easier to see the overplotted ones. Visually, does Mitch Hedberg’s joke appear to hold here?
...
```{r ggplot_for_north_carolina}
dn_lq %>%
  filter(state == "NC") %>%
ggplot(dn_lq, mapping = aes(x = longitude, y = latitude, color = establishment, alpha = 0.5)) +
  geom_point()
```

Mitch Hedberg’s joke does not appear to hold here. There are only a couple places where laquinta is next to Denny's. Most of the Denny's locations are from from La Quinta locations.

### Exercise 12
Now filter the data for observations in Texas only, and recreate the plot, with an appropriate alpha level. Visually, does Mitch Hedberg’s joke appear to hold here?
...

```{r ggplot_for_Texas}
dn_lq %>%
  filter(state == "TX") %>%
ggplot(dn_lq, mapping = aes(x = longitude, y = latitude, color = establishment, alpha = 0.3)) +
  geom_point()
```

Mitch Hedberg’s joke appears to hold here. Most of the Denny's locations seem to be right beside La Quinta hotels!



