---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 2"
author: "Insert your name here"
date: "Insert date here"
output: html_notebook
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("C:/Users/markf/Downloads/DS 241/states.csv")
```

### Exercise 1

Filter the Denny’s dataframe for Alaska (AK) and save the result as dn_ak. How many Denny’s locations are there in Alaska?

```{r create_dn_ak}
dn_ak <- dennys %>%
  filter(state == "AK")
nrow(dn_ak)
```
There are three Dennys locations in Alaska
### Exercise 2

Filter the La Quinta dataframe for Alaska (AK) and save the result as lq_ak. How many La Quinta locations are there in Alaska?

```{r create_lq_ak}
lq_ak <- laquinta %>%
  filter(state == "AK")
nrow(lq_ak)
```
There are 2 locations in Alaska

Next we’ll calculate the distance between all Denny’s and all La Quinta locations in Alaska. 

### Exercise 3
How many pairings are there between all Denny’s and all La Quinta locations in Alaska, i.e. how many distances do we need to calculate between the locations of these establishments in Alaska?

There are 6 pairings, which means we have to calculate 6 distances
...
```{r full_join_laquinta_and_dennys}
dn_lq_ak <- full_join(dn_ak, lq_ak, by = "state")
dn_lq_ak
```

### Exercise 4
How many observations are in the joined dn_lq_ak data frame? What are the names of the variables in this data frame.

There are 6 observations. The names of the variables are adress, city, state, zip code, longitude, latitude, establishment. The variables then repeat for laquinta
...

### Exercise 5
What function from the tidyverse do we use the add a new variable to a data frame while keeping the existing variables?

The function we would use is the mutate funciton
...

```{r haversine_function}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 = long1 * pi / 180
  lat1  = lat1  * pi / 180
  long2 = long2 * pi / 180
  lat2  = lat2  * pi / 180
  
  R = 6371 # Earth mean radius in km
  
  a = sin((lat2 - lat1)/2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1)/2)^2
  d = R * 2 * asin(sqrt(a))
  
  return( round(d,round) ) # distance in km
}
```


### Exercise 6
Calculate the distances between all pairs of Denny’s and La Quinta locations and save this variable as distance. Make sure to save this variable in THE dn_lq_ak data frame so that you can use it later.
...
```{r create_distance_variable_1}
dn_lq_ak <- dn_lq_ak %>% 
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 1))
```


### Exercise 7
Calculate the minimum distance between a Denny’s and La Quinta for each Denny’s location. To do so we group by Denny’s locations and calculate a new variable that stores the information for the minimum distance.
...

```{r find_minimum_distance_1}
dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(address.x) %>%
  summarise(closest = min(distance))
```


### Exercise 8
Describe the distribution of the distances Denny’s and the nearest La Quinta locations in Alaska. 
Also include an appripriate visualization and relevant summary statistics.
...

```{r dot_plot_visualisation}
ggplot(data = dn_lq_ak_mindist) + 
  geom_dotplot(mapping = aes(x = closest, y = address.x))
```
The distribution of the data is skewed left.
```{r summary_statistics}
summary(dn_lq_ak_mindist)
```


### Exercise 9
Repeat the same analysis for North Carolina:

(i) filter Denny’s and La Quinta Data Frames for NC,

(ii) join these data frames to get a completelist of all possible pairings,

(iii) calculate the distances between all possible pairings of Denny’s and La Quinta in NC, 

(iv) find the minimum distance between each Denny’s and La Quinta location,

(v) visualize and describe the distribution of these shortest distances using appropriate summary statistics.

(vi) create a function that allows steps i-v to be done quickly to any other state
...

complete step i
```{r create_dn_NC}
dn_nc <- dennys %>%
  filter(state == "NC")
nrow(dn_nc)
```
repreat for laquinta
```{r create_lq_nc}
lq_nc <- laquinta %>%
  filter(state == "NC")
nrow(lq_nc)
```

Complete step ii
```{r full_join_dn_nc_and_lq_nc}
dn_lq_nc <- full_join(dn_nc, lq_nc, by = "state")
dn_lq_nc
```

complete step iii
```{r create_distance_variable}
dn_lq_nc <- dn_lq_nc %>% 
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 1))
```

complete step iv
```{r find_minimum_distance}
dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(address.x) %>%
  summarise(closest = min(distance))
```

complete step v
```{r}
summary(dn_lq_nc_mindist)
ggplot(data = dn_lq_nc_mindist) + 
  geom_dotplot(mapping = aes(x = closest))
```

complete step vi
```{r create_function}
map_mindist <- function(which_state) {
  #step i
  x = which_state
  
  dn_x <- dennys %>%
  filter(state == x)

  lq_x <- laquinta %>%
  filter(state == x)
  
  dn_lq_x <- full_join(dn_x, lq_x, by = "state") %>% 
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 1))
  
  dn_lq_x_mindist <- dn_lq_x %>%
  group_by(address.x) %>%
  summarise(closest = min(distance))
  
#The below code will help you quickly print visualizations or summary statistics
  
# summary(dn_lq_x_mindist)
# ggplot(data = dn_lq_x_mindist) + 
#   geom_dotplot(mapping = aes(x = closest))
  
}
```

This code chunk may be disregarded. It was only used for development of the above funtion.
```{r testing}
# map_mindist("TX")
```

### Exercise 10
Repeat the same analysis for Texas.
...

```{r repeat_analysis_for_texas}
dn_lq_tx_mindist <- map_mindist("TX") 
summary(dn_lq_tx_mindist)
   ggplot(data = dn_lq_tx_mindist) + 
    geom_dotplot(mapping = aes(x = closest))
```


### Exercise 11
Repeat the same analysis for a state of your choosing, different than the ones we covered so far.
...

```{r repeat_analysis_for_ny}
dn_lq_ny_mindist <- map_mindist("NY") 
summary(dn_lq_ny_mindist)
   ggplot(data = dn_lq_ny_mindist) + 
    geom_dotplot(mapping = aes(x = closest))
```


### Exercise 12
Among the states you examined, where is Mitch Hedberg’s joke most likely to hold true? Explain your reasoning.

...
Among the states I examined, North Carolina, Texas, and New York, Mitch Hedberg's joke is most likely to hold true for Texas. The reason for this is that the mean minimum distance between a Denny's and the Nearest LaQuinta is the lowest. In Texas, the mean minimum distance is 5.792. For North Carolina and New York, the mean minimum distances are 65.44 and 33.577 respectively